Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-31 paper.tex

\subsection{Future applications: Cloud computing}
To explore the use of cloud computing services with VCF Zarr,
Genomics England provided access to Aggv2 data 
in VCF Zarr format within their secure
cloud environment. The VCF Zarr files for Chromosome 2
described in Table~\ref{tab-genomics-england-data}
were copied to a standard AWS S3 bucket.
Using this data we repeated the benchmarks above on the same subset
of 562,640 variants and 78,195 samples using the same
single-threaded sequential Zarr code.
Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed.
Creating the filter mask required 3.05min CPU, 11.15min elapsed.

While these benchmarks are comparable with results obtained on the
shared file-system in the Genomics England research environment,
the sequential single-threaded access patterns used do not map
well to modern cloud computing environments.
As each request to a cloud object store has a high latency, it
is important to have many requests running concurrently to achieve
good performance~\citep{durner2023exploiting}.
To explore the limits of performance we developed a simple prototype
Zarr implementation, designed to maximise concurrency using asynchronous
IO and multiple processes (see Methods for details).
Fig~\ref{fig-s3-throughput} shows the rate at which genotypes
can be decoded to RAM as well as rate at which we can compute
af-dist (using Numba to classify genotypes, as before)
as we increase the number of parallel processes on
an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
The maximum decompression rate of 47.4 GiB/s
occurs at 64 processes (each with 100 concurrent requests), 
with a corresponding network data rate of 121 MiB/s
Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
network access by measuring the rate at which we can download the chunks
into RAM (without decompressing) using the same parallel and asynchronous
architecture. Here, the maximum rate of 1.2 GiB/s for a c5.9xlarge instance
and 2.4 GiB/s for a c5n.9xlarge (high performance network) instance
was achieved with 32 processes. 
% Richard Durbin asked whether we were just using up all of a scarce 
% resource on this point.
While these measurements are simplistic and may not be an 
accurate predictor of sustained bandwidth, it is still clear that 
genotype data can be fetched
from S3 around an order of magnitude more quickly 
than it can be decompressed into RAM. 
We have also not measured
the achievable bandwidth while scaling out across many instances,
but given the ubiquity of cloud-based services and the scale 
at which they operate, it seems reasonable to assume that the claimed
terabits per second of aggregate bandwidth from
S3~\cite{s3-performance-optimising} are achievable.

@@

Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
calculation is performed at a maximum rate of 25.5 GiB/s using 128
processes. This level of throughput would complete af-dist for all
chromosomes for all samples (104 TiB of genotype data)
% timing: 722000000 * 79000 * 2bytes = 104 TiB
% 104TiB / 25.5GiB/s = 1.16 hours
% compute: cost of c5.9xlarge per hour * 1.16
% ==                             $4.08 * 1.16
% io: ((total # variants / variant chunk size) * (sample count/sample chunk size)) /1000 * cost for 1000 IOPS
% ==  (((722000000       / 10000)              * (79000       /1000))              /1000)* $0.0004
in 1.16 hours for a compute cost of \$4.73 and an IO cost of \$2.28.

Using this parallel architecture, the af-dist calculation for the target 
% 180.80 MB in 4503 files
% Total data: 90060.00 MB in 4503 chunks
region used for the single threaded benchmarks above ---172 MiB retrieved in 4503
chunks, 83.87 GiB of genotype data decompressed and processed---
% This gives an effective rate lower than  25.5G above:  83.87 / 4.15 = 20.2
was completed in 4.15s. A slightly slower effective throughput than in
Fig~\ref{fig-s3-throughput} as the short time to completion means that
the timing is sensitive to set-up and tear-down costs.
See \citep{durbin2020task,moore2021ome,gowan2022using} for further Zarr
benchmarks in cloud settings.

@@

Scientific datasets are increasingly overwhelming the classical
model of downloading and analysing locally, and are migrating to
centralised cloud repositories \citep{abernathey2021cloud,moore2021ome}.
The combination of Zarr's simple and cloud-friendly storage
of data chunks with state-of-the-art compression methods has
led to Zarr gaining significant traction in these settings.
Multiple petabyte-scale datasets are now stored using
Zarr~\cite[e.g.][]{gowan2022using, % weather model data
fahnestock2023mappin, % ITS_LIVE dataset
cmip6_dataset}
or under active consideration for migration~\citep{durbin2020task,abernathey2021opening}.
The Open GeoSpatial consortium has formally recognised Zarr as a community
standard~\cite{ogc_zarr2_standard}
and has formed
a new GeoZarr Standards Working Group to establish a Zarr encoding for
geospatial data~\cite{ogc_geozarr_news}.

>>>>

\subsection{Future applications: Cloud computing}
To explore the use of cloud computing services with VCF Zarr,
Genomics England provided access to aggV2 data 
in VCF Zarr format within their secure
cloud environment. The VCF Zarr files for Chromosome 2
described in Table~\ref{tab-genomics-england-data}
were copied to a standard AWS S3 bucket.
Using this data we repeated the benchmarks above on the same subset
of 562,640 variants and 78,195 samples using the same
single-threaded sequential Zarr code.
Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed.
Creating the filter mask required 3.05min CPU, 11.15min elapsed.

While these benchmarks are comparable with results obtained on the
shared file-system in the Genomics England research environment,
the sequential single-threaded access patterns used do not map
well to modern cloud computing environments.
As each request to a cloud object store has a high latency, it
is important to have many requests running concurrently to achieve
good performance~\citep{durner2023exploiting}.
To explore the limits of performance we developed a simple prototype
Zarr implementation, designed to maximise concurrency using asynchronous
IO and multiple processes (see Methods for details).
Fig~\ref{fig-s3-throughput} shows the rate at which genotypes
can be decoded to RAM along with the rate at which we can compute
af-dist (using Numba to classify genotypes, as before)
as we increase the number of parallel processes on
an EC2 c5.24xlarge instance with 96 vCPUs (48 physical).
The maximum decompression rate of 47.4 GiB/s
occurs at 64 processes (each with 100 concurrent requests), 
with a corresponding network data rate of 121 MiB/s.
Fig~\ref{fig-s3-network-throughput} explores the limitations imposed by
network access by measuring the rate at which we can download the chunks
into RAM (without decompressing) using the same parallel and asynchronous
architecture. Here, the maximum rate of 1.2 GiB/s for a c5.9xlarge instance
and 2.4 GiB/s for a c5n.9xlarge (high performance network) instance
was achieved with 32 processes. 
% Richard Durbin asked whether we were just using up all of a scarce 
% resource on this point.
While these measurements are simplistic and may not be an 
accurate predictor of sustained bandwidth, it is still clear that 
genotype data can be fetched
from S3 around an order of magnitude more quickly 
than it can be decompressed into RAM. 
We have also not measured
the achievable bandwidth while scaling out across many instances,
but given the ubiquity of cloud-based services and the scale 
at which they operate, it seems reasonable to assume that the claimed
terabits per second of aggregate bandwidth from
S3~\cite{s3-performance-optimising} are achievable.

@@

% >>> 78_195 * 722*10**6 * 2 / 1024**4
% 102.69430276821367
% >>> (78_195 * 722*10**6 * 2 / 1024**3) / 25.5
% 4123.881020966698
% >>> 4123.881020966698 / 360
% 1.1455
% >>> 4123.881020966698 / 60
% 68.73135034944497
% >>> 1.1455 * 4.848
% 5.553383999999999
Returning to Fig~\ref{fig-s3-throughput}, we can see that the af-dist
calculation is performed at a maximum rate of 25.5 GiB/s using 128
processes. Genome wide, the aggV2 dataset consists of
722 million variants for 78,195 samples, giving a total of around
103 TiB of genotype data. We could therefore process the entire
dataset at this maximum rate in about 69 minutes, with a compute
cost of \$5.55
(on-demand hourly rate of \$4.848 for a c5.24xlarge instance
in the eu-west-2 region, as-of 2025-01-31~\cite{ec2-pricing}).
AWS and other major clouds charge per object request (i.e. without
considering the number of bytes transferred) and so the IO
cost of this operation is determined by the number of chunks.
% >>> (10_000 * 1000 * 2) / 1024**2
% 19.073486328125
% >>> (78_195 * 722*10**6 * 2) / (10_000 * 1000 * 2)
% 5645679.0
% >>> (5645679.0 / 1000) *0.00042
% 2.3711851800000003
As each 10,000 $\times$ 1,000 chunk encodes 19 MiB of 
genotype data, we incur about 5.6 million object requests
to retrieve all chunks. 
At \$0.00042 per 1000 GET requests for S3 Standard
storage 
(eu-west-2 region as-of 2025-01-31~\cite{s3-pricing}), 
this comes to \$2.37.
Thus, it should be possible to perform af-dist and similar
calculations genome wide on aggV2 for less than \$10.
Assuming 50KiB per chunk (the average stored chunk size 
for chromosome 2 is 47.57 KiB) we would need a total
of 269 GiB to store the genotype data for the entire dataset.
% >>> (5645679.0 * 50 * 1024) / 1024**3
% 269.20695304870605
% 269.20695304870605 * 0.024 * 12
% 77.53160247802734
At \$0.024 per GiB per month for S3 Standard
(eu-west-2 region as-of 2025-01-31~\cite{s3-pricing})
this comes to a yearly storage cost of \$77.53
(note that this is just the genotype data, not including
other, much larger, fields 
listed in Table~\ref{tab-genomics-england-data}).

One of our benchmarks in the Genomics England case
study section was to compare to compare a straightforward
sequential Zarr implementation of af-dist with
\texttt{bcftools} 
on the genome slice corresponding to one of the original
106 VCF files for chromosome 2. There, Zarr's efficient data format
allows us to perform the same calculation in 4.25
% >>> 716.3 / 60
% 11.938333333333333
minutes vs almost 12 hours for VCF. 
Here, by adapting our code to the 
high-throughput, high-variance nature of cloud stores and 
using the natural chunk-based parallelism of Zarr
we reduce the elapsed time to 4.15 seconds.
These benchmarks illustrate the transformative 
potential of Zarr and cloud computing.

@@

Scientific datasets are increasingly overwhelming the classical
model of downloading and analysing locally, and are migrating to
centralised cloud repositories \citep{abernathey2021cloud,moore2021ome}.
The combination of Zarr's simple and cloud-friendly storage
of data chunks with state-of-the-art compression methods has
led to Zarr gaining significant traction in these settings
(see \citep{durbin2020task,moore2021ome,gowan2022using} 
Zarr cloud benchmarks).
Multiple petabyte-scale datasets are now stored using
Zarr~\cite[e.g.][]{gowan2022using, % weather model data
fahnestock2023mappin, % ITS_LIVE dataset
cmip6_dataset}
or under active consideration for migration~\citep{durbin2020task,abernathey2021opening}.
The Open GeoSpatial consortium has formally recognised Zarr as a community
standard~\cite{ogc_zarr2_standard}
and has formed
a new GeoZarr Standards Working Group to establish a Zarr encoding for
geospatial data~\cite{ogc_geozarr_news}.
